Linked Data Notifications: a resource-centric communication protocol

### 2017-03-08 index.xh2latex.tex
using proper LaTeX; restored figures and tables

\usepackage{listings}
%\lstset{numbers=left, captionpos=b, frame=single, basicstyle=\ttfamily, stringstyle=\small, breaklines=true, extendedchars=true, showstringspaces=true, numberbychapter=false}

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{hyperref}
\usepackage{enumitem}
\setlist[itemize,enumerate]{leftmargin=*}

@@

\par The {\em Linked Data Notifications (LDN)} protocol describes how servers (receivers) can receive messages pushed to them by applications (senders), as well as how other applications (consumers) may retrieve those messages. Any resource can advertise a receiving endpoint (Inbox) for notification messages. Messages are expressed in RDF, and can contain arbitrary data. It is not dependent on a complete implementation of LDP, but comprises an easy-to-implement subset. LDN is a \empty W3C Candidate Recommendation~\cite{ref-4}.

                                Overview of Linked Data Notifications

@@

                                        \begin{verbatim}{
  "@context": { "sioc": "http://rdfs.org/sioc/ns#" }
  "@id": "",
  "@type": "sioc:Comment",
  "sioc:content": "This is a great article!",
  "sioc:reply_of": { "@id": "http://example.org/article" },
  "sioc:created_at": { "@value": "2015-12-23T16:44:21Z" }
}\end{verbatim}
                                        A notification about a comment created by a user (JSON-LD).

                                        \begin{verbatim}@prefix as: <https://www.w3.org/ns/activitystreams#> .
@prefix cito: <http://purl.org/spar/cito/> .
<> a as:Announce
  as:object <https://linkedresearch.org/resources#r-903b83> ;
  as:target <http://csarven.ca/dokieli#architecture> .
<https://linkedresearch.org/resources#r-903b83>
  cito:citesAsPotentialReading
    <http://csarven.ca/linked-data-notifications#protocol> .\end{verbatim}
                                        An announcement of a specific citation relation between two entities (Turtle).

@@

                                LDN Implementations

                                        Implementation
                                        Class\empty *
                                        Description

                                        \begin{description}
  \item[*] Conformance classes: S – sender, C – consumer, R – receiver.\item[a] Implementations by the authors
    \end{description}

\par Source: \empty https://github.com/w3c/ldn/tree/master/implementations

                                        \empty CarbonLDP
                                        R
                                        Data storage platform (LDP)

                                        \empty dokielia
                                        S,C
                                        Clientside editor and annotator

                                        \empty errola
                                        S
                                        Generic message sending client

                                        \empty Fedora Commons
                                        R
                                        Open source repository platform (LDP)

                                        \empty IndieAnndroid
                                        R
                                        Personal blogging platform

                                        \empty Linked Edit Rules
                                        S
                                        Statistical dataset consistency checker

                                        \empty mayktsoa
                                        R
                                        Personal data store (LDP)

                                        \empty OnScreena
                                        C
                                        Notifications display client

                                        \empty pyldn
                                        R
                                        Standalone Inbox

                                        \empty RDF-LinkedData-Notifications
                                        R
                                        Standalone Inbox

                                        \empty slopha
                                        S,R
                                        Social publishing \& quantified self

                                        \empty Solid Words
                                        S
                                        Foreign language learning app

                                        \empty solid-client
                                        S
                                        Clientside library for LDP

                                        \empty solid-inbox
                                        C
                                        Clientside social message reader

                                        \empty solid-notifications
                                        S,C
                                        Clientside library for LDN

                                        \empty solid-server
                                        R
                                        Personal data storage server (LDP)

                                        \empty Virtuoso+\empty ODS Briefcase
                                        R,C
                                        Personal data storage server (LDP)

\par We highlight social scholarly communication use cases with \empty dokieli, a clientside editor for decentralised scientific article publishing, annotations and social interactions~\cite{ref-17}. dokieli uses LDN to send and consume notifications: When a reader comments on a fragment of text in an article, the application discovers the article’s Inbox and sends a notification about the annotation. dokieli also consumes notifications from this Inbox to fetch and display the annotation as marginalia (\empty figure 2). A reader can share a dokieli-enabled article with their contacts; dokieli discovers each contact’s Inbox and sends a notification there (\empty figure 3). When editing an article, the author can add a citation. If an Inbox is discovered in the cited article, dokieli sends a notification there to indicate what part of the article was cited by whom and where. dokieli-enabled articles also consume citation notifications to display these metrics for the author and other readers (\empty figure 4).

                                \empty Video of dokieli Web Annotation

                                \empty Video of dokieli Share

@@

                            \begin{itemize}
  \item If a {\em Like} activity is generated on the server, sloph uses the {\tt object} of the {\em Like} as the target for a notification. Since dokieli uses the same vocabulary for social interactions (\empty ActivityStreams 2.0~\cite{ref-18}), if the target is a dokieli article, this {\em Like} will be displayed (\empty figure 5).\item If the user publishes a blog post containing a link, which may be semantically annotated to indicate the reason for linking, sloph sends a notification to any Inbox discovered at that link.\item As a receiver, sloph accepts all incoming notifications, but holds for moderation (i.e. places behind access control) any that it cannot automatically verify refer to third-party content published on another domain. If an article written with dokieli publishes a citation of a blog post which advertises a sloph Inbox, sloph will fetch the article and verify whether the relation matches the contents of the notification before exposing the notification for re-use.
    \end{itemize}

\par \empty Linked Edit Rules and \empty Solid Words are specialised senders. Linked Edit Rules checks the consistency of statistical datasets against structured constraints, and delivers the consistency report as a notification to the user. Solid Words is a clientside game for learning new words in a foreign language; it delivers the player’s score for each round to their Inbox. \empty OnScreen is a (crude) generic consumer; as such, it can display notifications sent by both of the aforementioned senders (\empty figure 6).

                                A {\em Like} notification created by sloph, displayed by dokieli.

                                A: Solid Words (a sender), B: Linked Edit Rules (a sender), C: OnScreen (a consumer) displaying notifications sent by A and B.

@@

                                        Comparison of notification mechanisms

                                                Mechanism
                                                T1
                                                T2
                                                T3
                                                R1
                                                R2
                                                R3
                                                R4-A
                                                R4-B
                                                R4-Cp
                                                R4-Cv
                                                R4-Co
                                                R5

                                                    \begin{description}
  \item[T1] Notification type\item[T2] Delivery method\item[T3] Dependencies\item[R1] Modularity (application classes: S Sender, R Receiver, C Consumer, U Subscriber/User)\item[R2] Reusability\item[R3] Persistence - required? how?\item[R4-A] Target representation\item[R4-B] Notification body\item[R4-Cp] Payload processing required?\item[R4-Cv] Verification - required? how?\item[R4-Co] Requirements for referenced resources?\item[R5] Subscription
    \end{description}

                                                    \begin{description}
  \item[–] not applicable, out of scope\item[/] not specified, in scope\item[X] explicitly disallowed\item[app] application specific decision\item[!] required ({\em MUST})\item[+] recommended ({\em SHOULD})\item[O] optional ({\em MAY})\item[PuSH] PubSubHubbub
    \end{description}

                                                    \begin{description}
  \item[h] HTML recommended\item[j] Alternate RDF formats can be negotiated\item[k] {\tt source} and {\tt target} key–value pairs is required\item[q] Provenance records with \empty PROV Ontology\item[r] RDF representation recommended\item[ra] SPARQL results transformed to RSS/Atom\item[s] \empty Sitemaps\item[t] Described in an RDF store or dataset
    \end{description}

                                                Semantic Pingback
                                                Linkback
                                                POST
                                                RDF
                                                S R
                                                /
                                                /
                                                Anyr
                                                form urlencodedk
                                                !
                                                ! parse source
                                                Anyr
                                                X

                                                Webmention
                                                Linkback
                                                POST
                                                HTML
                                                S R
                                                –
                                                –
                                                Anyh
                                                form urlencodedk
                                                !
                                                ! parse source
                                                Anyh
                                                X

                                                Provenance Pingback
                                                Linkback
                                                POST
                                                RDF
                                                S R
                                                /
                                                /
                                                /
                                                URI list
                                                /
                                                /
                                                RDFq
                                                X

                                                DSNotify
                                                Fat ping
                                                POST, PUT
                                                XML, PuSH
                                                S U
                                                /
                                                –
                                                –
                                                XML
                                                /
                                                –
                                                RDFt
                                                !

                                                sparqlPuSH
                                                Fat ping
                                                POST
                                                XML, SPARQL, PuSH
                                                S U
                                                –
                                                –
                                                –
                                                XMLra
                                                /
                                                –
                                                RDFt
                                                !

                                                ResourceSync
                                                Fat ping
                                                POST
                                                XML, PuSH
                                                S U
                                                /
                                                –
                                                –
                                                XMLs
                                                /
                                                –
                                                ?
                                                !

                                                Linked Data Notifications
                                                Fat ping
                                                POST
                                                JSON-LD
                                                S R C
                                                !
                                                ! URI
                                                Any
                                                JSON-LDj
                                                + app
                                                + app
                                                –
                                                O app

@@

>>>>

\usepackage{listings}
\lstset{numbers=left, captionpos=b, frame=single, basicstyle=\ttfamily, stringstyle=\small, breaklines=true, extendedchars=true, showstringspaces=true, numberbychapter=false}

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{ctable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{enumitem}
\setlist[itemize,enumerate]{leftmargin=*}

@@

\par The {\em Linked Data Notifications (LDN)} protocol describes how servers (receivers) can receive messages pushed to them by applications (senders), as well as how other applications (consumers) may retrieve those messages. Any resource can advertise a receiving endpoint (Inbox) for notification messages. Messages are expressed in RDF, and can contain arbitrary data. It is not dependent on a complete implementation of LDP, but comprises an easy-to-implement subset.

\begin{figure}
  \centering
  \includegraphics[trim=.2cm 10cm .5cm 10cm,clip,width=.6\textwidth]{media/images/{linked-data-notifications-overview.svg}.pdf}
  \caption{Overview of Linked Data Notifications}
  \label{fig:overview}
\end{figure}

@@

\begin{lstlisting}
{ "@context": { "sioc": "http://rdfs.org/sioc/ns#" }
  "@id": "",
  "@type": "sioc:Comment",
  "sioc:content": "This is a great article!",
  "sioc:reply_of": { "@id": "http://example.org/article" },
  "sioc:created_at": { "@value": "2015-12-23T16:44:21Z" } }
\end{lstlisting}
                                        A notification about a comment created by a user (JSON-LD).

\begin{lstlisting}
@prefix as: <https://www.w3.org/ns/activitystreams#> .
@prefix cito: <http://purl.org/spar/cito/> .
<> a as:Announce
  as:object <https://linkedresearch.org/resources#r-903b83> ;
  as:target <http://csarven.ca/dokieli#architecture> .
<https://linkedresearch.org/resources#r-903b83>
  cito:citesAsPotentialReading
    <http://csarven.ca/linked-data-notifications#protocol> .
\end{lstlisting}
                                        An announcement of a specific citation relation between two entities (Turtle).

@@

\ctable[
  caption={LDN Implementations},
  label = {tab:implementations}
]{lll}{
  \tnote[*]{Conformance classes: S – sender, C – consumer, R – receiver.}
  \tnote[a]{Implementations by the authors}
  \tnote[]{Source: \url{https://github.com/w3c/ldn/tree/master/implementations}}
}{\FL
  Implementation & Class\tmark[*] & Description
  \ML
                                        CarbonLDP &
                                        R &
                                        Data storage platform (LDP) \NN

                                        dokieli\tmark[a] &
                                        S,C &
                                        Clientside editor and annotator \NN

                                        errol\tmark[a] &
                                        S &
                                        Generic message sending client \NN

                                        Fedora Commons &
                                        R &
                                        Open source repository platform (LDP) \NN

                                        IndieAnndroid &
                                        R &
                                        Personal blogging platform \NN

                                        Linked Edit Rules &
                                        S &
                                        Statistical dataset consistency checker \NN

                                        mayktso\tmark[a] &
                                        R &
                                        Personal data store (LDP) \NN

                                        OnScreen\tmark[a] &
                                        C &
                                        Notifications display client \NN

                                        pyldn &
                                        R &
                                        Standalone Inbox \NN

                                        RDF-LinkedData-Notifications &
                                        R &
                                        Standalone Inbox \NN

                                        sloph\tmark[a] &
                                        S,R &
                                        Social publishing \& quantified self \NN

                                        Solid Words &
                                        S &
                                        Foreign language learning app \NN

                                        solid-client &
                                        S &
                                        Clientside library for LDP \NN

                                        solid-inbox &
                                        C &
                                        Clientside social message reader \NN

                                        solid-notifications &
                                        S,C &
                                        Clientside library for LDN \NN

                                        solid-server &
                                        R &
                                        Personal data storage server (LDP) \NN

                                        Virtuoso+ ODS Briefcase &
                                        R,C &
                                        Personal data storage server (LDP) \LL
}

\par We highlight social scholarly communication use cases with \empty dokieli, a clientside editor for decentralised scientific article publishing, annotations and social interactions~\cite{ref-17}. dokieli uses LDN to send and consume notifications: When a reader comments on a fragment of text in an article, the application discovers the article’s Inbox and sends a notification about the annotation. dokieli also consumes notifications from this Inbox to fetch and display the annotation as marginalia (figure~\ref{fig:video-annotation}). A reader can share a dokieli-enabled article with their contacts; dokieli discovers each contact’s Inbox and sends a notification there (figure~\ref{fig:video-share}). When editing an article, the author can add a citation. If an Inbox is discovered in the cited article, dokieli sends a notification there to indicate what part of the article was cited by whom and where. dokieli-enabled articles also consume citation notifications to display these metrics for the author and other readers (figure~\ref{fig:like}).

\begin{figure}
  \begin{minipage}[b]{.49\textwidth}
    \includegraphics[width=\textwidth]{media/images/dokieli-annotation}
    \caption{Video of dokieli Web Annotation}
    \label{fig:video-annotation}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{.49\textwidth}
    \includegraphics[width=\textwidth]{media/images/dokieli-share}
    \caption{Video of dokieli Share}
    \label{fig:video-share}
  \end{minipage}
\end{figure}

@@

                            \begin{itemize}
  \item If a {\em Like} activity is generated on the server, sloph uses the {\tt object} of the {\em Like} as the target for a notification. Since dokieli uses the same vocabulary for social interactions (\empty ActivityStreams 2.0~\cite{ref-18}), if the target is a dokieli article, this {\em Like} will be displayed (figure~\ref{fig:like}).\item If the user publishes a blog post containing a link, which may be semantically annotated to indicate the reason for linking, sloph sends a notification to any Inbox discovered at that link.\item As a receiver, sloph accepts all incoming notifications, but holds for moderation (i.e. places behind access control) any that it cannot automatically verify refer to third-party content published on another domain. If an article written with dokieli publishes a citation of a blog post which advertises a sloph Inbox, sloph will fetch the article and verify whether the relation matches the contents of the notification before exposing the notification for re-use.
    \end{itemize}

\par \empty Linked Edit Rules and \empty Solid Words are specialised senders. Linked Edit Rules checks the consistency of statistical datasets against structured constraints, and delivers the consistency report as a notification to the user. Solid Words is a clientside game for learning new words in a foreign language; it delivers the player’s score for each round to their Inbox. \empty OnScreen is a (crude) generic consumer; as such, it can display notifications sent by both of the aforementioned senders (figure~\ref{fig:notifications}).

\begin{figure}
  \includegraphics[width=\textwidth]{media/images/screenshot-ldn-sloph-dokieli}
  \caption{A {\em Like} notification created by sloph, displayed by dokieli.}
  \label{fig:like}

  \includegraphics[width=\textwidth]{media/images/screenshot-ldn-senders}
  \caption{A: Solid Words (a sender), B: Linked Edit Rules (a sender), C: OnScreen (a consumer) displaying notifications sent by A and B.}
  \label{fig:notifications}
\end{figure}

@@

\begin{table}
  \caption{Comparison of notification mechanisms}
  \label{tab:comparison}
  \footnotesize
  \begin{tabular}{% Mechanism
    p{.15\textwidth}
    % T1–T3
    p{.07\textwidth}
    p{.07\textwidth}
    p{.09\textwidth}
    % R1–R3
    p{.03\textwidth}
    p{.03\textwidth}
    p{.05\textwidth}
    % R4-A–R5
    p{.06\textwidth}
    p{.14\textwidth}
    p{.045\textwidth}
    p{.06\textwidth}
    p{.06\textwidth}
    p{.05\textwidth}
    }\FL
                                                Mechanism&
                                                T1&
                                                T2&
                                                T3&
                                                R1&
                                                R2&
                                                R3&
                                                R4-A&
                                                R4-B&
                                                R4-C\textsuperscript{p}&
                                                R4-C\textsuperscript{v}&
                                                R4-C\textsuperscript{o}&
                                                R5\ML
                                                Semantic\newline Pingback&
                                                Link-\newline back&
                                                POST&
                                                RDF&
                                                S\newline R&
                                                /&
                                                /&
                                                Any\textsuperscript{r}&
                                                form\newline urlencoded\textsuperscript{k}&
                                                !&
                                                !\newline parse\newline source&
                                                Any\textsuperscript{r}&
                                                X\NN
                                                Webmention&
                                                Link-\newline back&
                                                POST&
                                                HTML&
                                                S\newline R&
                                                –&
                                                –&
                                                Any\textsuperscript{h}&
                                                form\newline urlencoded\textsuperscript{k}&
                                                !&
                                                !\newline parse\newline source&
                                                Any\textsuperscript{h}&
                                                X\NN
                                                Provenance\newline Pingback&
                                                Link-\newline back&
                                                POST&
                                                RDF&
                                                S\newline R&
                                                /&
                                                /&
                                                /&
                                                URI list&
                                                /&
                                                /&
                                                RDF\textsuperscript{q}&
                                                X\NN
                                                DSNotify&
                                                Fat\newline ping&
                                                POST,\newline PUT&
                                                XML,\newline PuSH&
                                                S\newline U&
                                                /&
                                                –&
                                                –&
                                                XML&
                                                /&
                                                –&
                                                RDF\textsuperscript{t}&
                                                !\NN
                                                sparqlPuSH&
                                                Fat\newline ping&
                                                POST&
                                                XML,\newline SPARQL,\newline PuSH&
                                                S\newline U&
                                                –&
                                                –&
                                                –&
                                                XML\textsuperscript{ra}&
                                                /&
                                                –&
                                                RDF\textsuperscript{t}&
                                                !\NN
                                                ResourceSync&
                                                Fat\newline ping&
                                                POST&
                                                XML,\newline PuSH&
                                                S\newline U&
                                                /&
                                                –&
                                                –&
                                                XML\textsuperscript{s}&
                                                /&
                                                –&
                                                ?&
                                                !\NN
                                                Linked Data\newline Notifications&
                                                Fat\newline ping&
                                                POST&
                                                JSON-LD&
                                                S\newline R\newline C&
                                                !&
                                                !\newline URI&
                                                Any&
                                                JSON-LD\textsuperscript{j}&
                                                + app&
                                                + app&
                                                –&
                                                O\newline app\LL
  \end{tabular}\\
  \begin{tabularx}{1.0\linewidth}{XX}\FL
    \textbf{T1}: Notification type&
    \textbf{R4-A}: Target representation\NN
    \textbf{T2}: Delivery method&
    \textbf{R4-B}: Notification body\NN
    \textbf{T3}: Dependencies&
    \textbf{R4-C\textsuperscript{p}}: Payload processing required?\NN
    \textbf{R1}: Modularity (application classes: S Sender, R Receiver, C Consumer, U Subscriber/User)&
    \textbf{R4-C\textsuperscript{v}}: Verification - required? how?\NN
    \textbf{R2}: Reusability&
    \textbf{R4-C\textsuperscript{o}}: Requirements for referenced resources?\NN
    \textbf{R3}: Persistence - required? how?&
    \textbf{R5}: Subscription\ML
    \textbf{–}: not applicable, out of scope&
    \textbf{!}: required ({\em MUST})\NN
    \textbf{/}: not specified, in scope&
    \textbf{+}: recommended ({\em SHOULD})\NN
    \textbf{X}: explicitly disallowed&
    \textbf{O}: optional ({\em MAY})\NN
    \textbf{app}: application specific decision&
    \textbf{PuSH}: PubSubHubbub\ML
    \textbf{\textsuperscript{h}}: HTML recommended&
    \textbf{\textsuperscript{r}}: RDF representation recommended\NN
    \textbf{\textsuperscript{j}}: Alternate RDF formats can be negotiated&
    \textbf{\textsuperscript{ra}}: SPARQL results transformed to RSS/Atom\NN
    \textbf{\textsuperscript{k}}: {\tt source} and {\tt target} key–value pairs is required&
    \textbf{\textsuperscript{s}}: \empty Sitemaps\NN
    \textbf{\textsuperscript{q}}: Provenance records with \empty PROV Ontology&
    \textbf{\textsuperscript{t}}: Described in an RDF store or dataset\LL
  \end{tabularx}
\end{table}

@@

\paragraph{Acknowledgements:}
This research was supported in part by Massachusetts Institute of Technology and Qatar Computing Research Institute, HBKU through the Crosscloud project from 2015-10 to 2016-09, and by the DFG project “Opening Scholarly Communication in Social Sciences” (grant agreement AU 340/9-1).
